KIN: A method to infer relatedness from low-coverage ancient DNA

### 2022-08-18 paper_overleaf/main.tex
Updates from Overleaf

However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from an out-of-sample reference panel to calculate the probability of IBD given IBS. There are several methods that incproporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022,browning_fast_2011,li_accurate_2014}.

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings without and with runs of homozygosity (ROH). In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes. The bottom row shows the expected proportion of differences ($p$) along the chromosome. In both cases, $p$ can take values of $p_0$, $p_1$ or $p_2$, which are the expected proportion of differences in IBD states 0,1 and 2, respectively.  }
    \label{fig0:schematic}
\end{figure}

@@

\subsection{How our method works}
Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM)-based approach to estimate genetic kinship and IBD from low-coverage ancient DNA data. KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child relationships. KIN is also able to take into account ROH and contamination, and is not sensitive to SNP ascertainment. We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.

@@

\paragraph{Beta Binomial Model:}
The Binomial Model laid out above assumes that sites are independent. However, due to genetic linkage, nearby sites will be correlated. As a result, we empirically find that the data often has considerably higher variance than would be expected from a binomial model (Fig. S2). We take this into account by adding an over-dispersion parameter $\delta$. Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter.

@@

This parameterization of the beta distribution in terms of expected value $p$ and over-dispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parameterization in terms of $\alpha$ and $\beta$. We use this equation even if preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.

@@

\begin{align}\label{eq:10}
    log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\delta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \delta) + log P(\mathbf{Y}|\Theta)\nonumber\\
 &= [\sum_{w} log P(\Delta_w|Y_w, M, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, A)] + P(Y_0| \pi),
\end{align}

@@

We assume that contamination rates in both individuals are known and small ($<5\%$), and set $C_{ij} \approx C_i + C_j$, where $C_i, C_j$ are the contamination estimates from the two individuals. We also assume the divergence $\phi$ between our 
target population and the putative contaminant popoulation is known. With probability $C_{ij}$, a comparison between two random reads from the pair of tested individual will contain a contaminant read, and thus contain a difference with probability $\phi$, and with probability $1 - C_{ij}$ it will be between endogenous ones. The comparisons between two contaminant reads are ignored, since we assume $C_{ij}$ to be small.

@@

For any particular comparison showing a difference $D$, we calculate the probability of the event $E$ that it is between endogenous reads as 
\begin{equation}
    P(E|D) = \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}.
\end{equation}

@@

\begin{table}[h!]
\caption{\label{tab:Table 2}Model performance for ROH prediction. Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination, R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
    \hline
    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RA(bias) & RA(var) & RC(bias) & RC(var)\\
    \hline
    4x & 0.007 & 0.006 & 0.081 & 0.036 & 0.083 & 0.036 & 0.091 & 0.039\\
    \hline
    0.2x & 0.003 & 0.003 & 0.068 & 0.03 & 0.068 & 0.026 & 0.076 & 0.029\\
    \hline
    0.1x & 0.003 & 0.002 & 0.049 & 0.021 & 0.05 & 0.021 & 0.057 & 0.022\\
    \hline
\end{tabular}
\label{table1}
\end{table}

\paragraph{IBD prediction:}
We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatdness and coverages (Fig. \ref{fig4:IBDstate_accuracy}). The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.836, depending on relatedness. However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.563 at 0.03x. The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage. We find that adding contamination, ascertainment bias in our simulations has little effect on IBD prediction. Adding ROH to the simulations reduces the IBD prediction in two cases: the accuracy for second degree is decreased by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig. S9). We see this adverse effect of ROH in case of siblings, perhaps because only in case of siblings we do not introduce long ROH directly, but long and short ROH may be present through parents (see section \ref{method}). Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$). However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see Fig. \ref{fig3:Comparison_READ_KIN})

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
    \centering
    \caption{Evaluation of IBD estimation at different coverages. y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis. The error bars are drawn at 1 standard deviation from the mean. Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
    \label{fig4:IBDstate_accuracy}
\end{figure}

\paragraph{Relatedness classification:}
We evaluated the classification accuracy of KIN (cutoff: 1 log-likelihood unit) and compared it to that of READ (cutoff: 1 standard deviation)) (Fig. \ref{fig3:Comparison_READ_KIN}). We first describe the results for relatedness cases detectable by READ, viz. identical, $1^{st}$ degree, $2^{nd}$ degree and unrelated. In this case we show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data. The true positive rate is above 0.97 for both KIN and READ, while the false negative rate is below 0.02.  One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.94 to 0.86, compared to range of 0.91 to 0.74 for READ. The false positive rate in this case remains below 0.02 for both methods. We also find that READ is strongly impacted by contamination ($<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.53 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.04 and 0.17 for unrelated individuals and 2nd degree relatives, respectively. By contrast, the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control. For simulations with ROH, we find that KIN also outperforms READ for 1st and 2nd degree relatives, although both the true and false positive rates increased for both methods compared to the control for 2nd degree relatives. The increases in false positives is likely due to ROH making related individuals more similar. Finally, we also found that KIN has good power to detect relatedness cases that are not detectable by READ, i.e. parent-child, siblings and, to a lower level, 3rd degree relatives (Fig. \ref{fig3:Comparison_READ_KIN}, for details see table S4).

@@

\paragraph{Chagyrskaya and Okladnikov Neandertals:}
To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya and Okladnikov Caves in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}. This dataset contains genetic data from a total of 16 skeletal remains that likely belong to contemporary Neandertals who occupied the Chagyrskaya and Okladnikov caves between 59 and 51 kya, and atleast 44 kya respectively \cite{laurits_skov_genetic_nodate}. DNA extracted from each of these remains were captured with an array targeting variable sites identified in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}. This genetic data has  low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and DNA contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}. We focused our analysis on the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} as done by the authors for the relatedness analysis \cite{laurits_skov_genetic_nodate}. Our results for pairwise relatedness for these individuals are shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three specimens from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). Further, we identify Chagyrskaya17 and Chagyrskaya60 as 3rd-degree relatives (table S1).

@@

We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech Valley \cite{mittnik_kinship-based_2019}. We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin. We found that KIN was able to confidently classify $85\%$ of the 6903 possible comparisons ($\Delta LL<1$), READ $94\%$ ($sd<1$) and lcMLkin 53\% of pairs (\cite{mittnik_kinship-based_2019}).

@@

For less than third-degree relations, only eight cases are classified differently between KIN and READ. lcMLkin matches KIN's prediction in two cases, and does not have enough data in 5 cases. In the last case, all three methods differ (table S3, Fig. S11). Finally, there are three disagreements between KIN and lcMLkin in classification of parent-child versus siblings (READ predicts first degree for all three pairs). We plotted the pairwise differences for these three pairs in Fig. S12, and found that the proportion of differences along the genome aligns with the prediction of KIN.

@@

We show that KIN reliably detects IBD tracts for upto 0.1x coverage even in presence of contamination and ascertainment bias, and the accuracy for IBD detection reduces with lower coverages. Adding ROH to the simulations adversely affects IBD prediction in siblings (Fig. S9), but this does not affect the power to correctly classify siblings (Fig. \ref{fig3:Comparison_READ_KIN}).

@@

One limitation of our approach is that it assumes a single population. In case of a highly structured population, KIN may show inaccurate inference of $p_0$ causing inaccurate relatedness inference. Also, our method makes the assumption, that the median pairwise genetic difference in the population reflects the population diversity $p_0$, which fails if almost all individuals in the dataset are related. We may get around this problem by using an estimate of $p_0$, calculated from known a pair of unrelated individuals from same population, or another population with similar diversity. The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.

While we have focused on the application of KIN on ancient human samples, the model is not tied to this system. Assuming we know the recombination rate, and hence can estimate the transition matrix (see section \ref{method}), KIN can be widely applied to any diploid species. In addition, the output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio. This makes KIN easy to automatize for large datasets. To make application of KIN user-friendly, we provide a python package (KINgaroo) to create input files for KIN from processed bamfiles, while optionally estimating ROH, and correcting for contamination estimates.

@@

\section{Acknowledgements}
We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Laurits Skov, Leonardo N. M. Iasi, Alba Bossoms Mesa, Arev P. Sümer, Zuzana Hofmanová, Guido Alberto Gnecchi Ruscone, Ke Wang and Luca Traverso for helpful comments and discussions. This work was supported by the Max Planck Society and the European Research Council (Grant No.: 694707) to Svante Pääbo.

@@

\section{Data and material availability}
An open-source implementation of KIN and KINgaroo in python along with a toy example dataset, and the scripts to generate our test simulations are available on GitHub \url{https://github.com/DivyaratanPopli/Kinship_Inference}. We have deposited the version of software used in the manuscript on Zenodo (). The analysed datasets from Bronze Age Lech Valley were generated in a previous study \cite{mittnik_kinship-based_2019}. Chagyrskaya and Okladnikov dataset were generated from a study currently in press \cite{laurits_skov_genetic_nodate}, and the data will be uploaded to European Nucleotide Archive upon publication.

>>>>

However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from an out-of-sample reference panel to calculate the probability of IBD given IBS. There are several methods that incorporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022,browning_fast_2011,li_accurate_2014}.

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib_rev.png}
    \centering
    \caption{IBD sharing between siblings without and with runs of homozygosity (ROH). In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes. The bottom row shows the expected proportion of differences ($p$) along the chromosome. In both cases, $p$ can take values of $p_0$, $p_1$ or $p_2$, which are the expected proportion of differences in IBD states 0,1 and 2, respectively.  }
    \label{fig0:schematic}
\end{figure}

@@

\subsection{How our method works}
Here, we present KIN (Kinship Inference), a Hidden Markov Model (HMM)-based approach to estimate genetic kinship and IBD from low-coverage ancient DNA data. KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child relationships. KIN is also able to take into account ROH and contamination, and is not sensitive to SNP ascertainment. We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.

@@

\paragraph{Beta Binomial Model:}
We empirically find that the data often has considerably higher variance than would be expected from a binomial model (Fig. S2). We take this into account by adding an over-dispersion parameter $\delta$. Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter.

@@

This parameterization of the beta distribution in terms of expected value $p$ and over-dispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parameterization in terms of $\alpha$ and $\beta$. We use this equation even if preprocessing steps (see equation \ref{eq:x} and section \ref{contam}) result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.

@@

\begin{align}\label{eq:10}
    log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\delta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \delta) + log P(\mathbf{Y}|\Theta)\nonumber\\
 &= [\sum_{w} log P(\Delta_w|Y_w, M_w, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, A)] + P(Y_0| \pi),
\end{align}

@@

We assume that contamination rates in both individuals are known and small ($<5\%$), and set $C_{ij} = C_i + C_j$, where $C_i, C_j$ are the contamination estimates from the two individuals. We also assume the divergence $\phi$ between our 
target population and the putative contaminant popoulation is known. With probability $C_{ij}$, a comparison between two random reads from the pair of tested individual will contain a contaminant read, and thus contain a difference with probability $\phi$, and with probability $1 - C_{ij}$ it will be between endogenous ones. The comparisons between two contaminant reads are ignored, since we assume $C_{ij}$ to be small.

@@

For any particular comparison showing a difference $D$ (not to be confused with number of differences $D_w$), we calculate the probability of the event $E$ that it is between endogenous reads as 
\begin{equation}
    P(E|D) = \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}.
\end{equation}

@@

\begin{table}[h!]
\caption{\label{tab:Table 2}Model performance for ROH prediction. Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination), R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
    \hline
    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RA(bias) & RA(var) & RC(bias) & RC(var)\\
    \hline
    4x & 0.007 & 0.006 & 0.081 & 0.036 & 0.083 & 0.036 & 0.091 & 0.039\\
    \hline
    0.2x & 0.003 & 0.003 & 0.068 & 0.03 & 0.068 & 0.026 & 0.076 & 0.029\\
    \hline
    0.1x & 0.003 & 0.002 & 0.049 & 0.021 & 0.05 & 0.021 & 0.057 & 0.022\\
    \hline
\end{tabular}
\label{table1}
\end{table}

\paragraph{IBD prediction:}
We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatedness and coverages (Fig. \ref{fig4:IBDstate_accuracy}). The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.836, depending on relatedness. However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.563 at 0.03x. The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage. We find that adding contamination, ascertainment bias in our simulations has little effect on IBD prediction. Adding ROH to the simulations reduces the IBD prediction in two cases: the accuracy for second degree is decreased by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig. S9). We see this adverse effect of ROH in case of siblings, perhaps because only in case of siblings we do not introduce long ROH directly, but long and short ROH may be present through parents (see section \ref{method}). Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$). However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see Fig. \ref{fig3:Comparison_READ_KIN})

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
    \centering
    \caption{Evaluation of IBD estimation at different coverages. y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis. The error bars are drawn at 1 standard deviation from the mean. Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other. Accuracy is defined as the sum of true positives and true negatives divided by the total number of windows. We consider a prediction is true if it matches the IBD state at the central position of the window.}
    \label{fig4:IBDstate_accuracy}
\end{figure}

\paragraph{Relatedness classification:}
We evaluated the classification accuracy of KIN (cutoff: 1 log-likelihood unit) and compared it to that of READ (cutoff: 1 standard deviation)) (Fig. \ref{fig3:Comparison_READ_KIN}). We first describe the results for relatedness cases detectable by READ, viz. identical, $1^{st}$ degree, $2^{nd}$ degree and unrelated. In this case we show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data. The true positive rate is above 0.97 for both KIN and READ, while the false negative rate is below 0.02.  One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.94 to 0.86, compared to range of 0.91 to 0.74 for READ. The false positive rate in this case remains below 0.02 for both methods. We also find that READ is strongly impacted by contamination (proportion of contamination$<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.53 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.04 and 0.17 for unrelated individuals and 2nd degree relatives, respectively. By contrast, the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control. For simulations with ROH, we find that KIN also outperforms READ for 1st and 2nd degree relatives, although both the true and false positive rates increased for both methods compared to the control for 2nd degree relatives. The increases in false positives is likely due to ROH making related individuals more similar. Finally, we also found that KIN has good power to detect relatedness cases that are not detectable by READ, i.e. parent-child, siblings and, to a lower level, 3rd degree relatives (Fig. \ref{fig3:Comparison_READ_KIN}, for details see table S4).

@@

\paragraph{Chagyrskaya and Okladnikov Neandertals:}
To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya and Okladnikov Caves in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}. This dataset contains genetic data from a total of 16 skeletal remains that likely belong to contemporary Neandertals who occupied the Chagyrskaya and Okladnikov caves between 59 and 51 kya, and at least 44 kya respectively \cite{laurits_skov_genetic_nodate}. DNA extracted from each of these remains were captured with an array targeting variable sites identified in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}. This genetic data has  low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and DNA contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}. We focused our analysis on the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} as done by the authors for the relatedness analysis \cite{laurits_skov_genetic_nodate}. Our results for pairwise relatedness for these individuals are shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three specimens from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). Further, we identify Chagyrskaya17 and Chagyrskaya60 as 3rd-degree relatives (table S1).

@@

We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech Valley \cite{mittnik_kinship-based_2019}. We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin. We found that KIN was able to confidently classify $85\%$ of the 6903 possible comparisons ($\Delta LL>1$), READ $94\%$ ($s.d.>1$) and lcMLkin 53\% of pairs (\cite{mittnik_kinship-based_2019}).

@@

For less than third-degree relations, only eight cases are classified differently between KIN and READ. lcMLkin matches KIN's prediction in two cases, and does not have enough data in 5 cases. In the last case, all three methods differ (table S3, Fig. S11), and the true relatedness is unresolved in this case. Finally, there are three disagreements between KIN and lcMLkin in classification of parent-child versus siblings (READ predicts first degree for all three pairs). We plotted the pairwise differences for these three pairs in Fig. S12, and found that the proportion of differences along the genome aligns with the prediction of KIN.

@@

We show that KIN reliably detects IBD tracts for up to 0.1x coverage even in presence of contamination and ascertainment bias, and the accuracy for IBD detection reduces with lower coverages. Adding ROH to the simulations adversely affects IBD prediction in siblings (Fig. S9), but this does not affect the power to correctly classify siblings (Fig. \ref{fig3:Comparison_READ_KIN}).

@@

One limitation of our approach is that it assumes a single population. In case of a highly structured population, KIN may show inaccurate inference of $p_0$ causing inaccurate relatedness inference. Also, our method makes the assumption, that the median pairwise genetic difference in the population reflects the population diversity $p_0$, which fails if almost all individuals in the dataset are related. We may get around this problem by using an estimate of $p_0$, calculated from known a pair of unrelated individuals from same population, or another population with similar diversity. We provide the user with an option to provide an estimate of $p_0$. The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.

While we have focused on the application of KIN on ancient human samples, the model is not tied to this system. Assuming we know the recombination rate, and hence can estimate the transition matrix (see section \ref{method}), KIN can be widely applied to any diploid species. In addition, the output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio. This makes KIN easy to automatize for large datasets. To make application of KIN user-friendly, we provide a python package (KINgaroo) to create input files for KIN from processed bam files, while optionally estimating ROH, and correcting for contamination estimates.

@@

\section{Acknowledgements}
We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Zbigniew Jędrzejewski-Szmek, Laurits Skov, Leonardo N. M. Iasi, Alba Bossoms Mesa, Arev P. Sümer, Zuzana Hofmanová, Guido Alberto Gnecchi Ruscone, Ke Wang and Luca Traverso for helpful comments and discussions. This work was supported by the Max Planck Society and the European Research Council (Grant No.: 694707) to Svante Pääbo.

@@

\section{Data and material availability}
An open-source implementation of KIN and KINgaroo in python along with a toy example dataset, and the scripts to generate our test simulations are available on GitHub \cite{divyaratan_popli_httpsgithubcomdivyaratanpoplikinship_inferencereleasestagv312_2022}. We have deposited the version of the software used in the manuscript, along with the above mentioned files on Zenodo \cite{divyaratan_popli_httpsdoiorg105281zenodo6561015_nodate}. The analysed datasets from Bronze Age Lech Valley were generated in a previous study (ENA, accession no. PRJEB34400). Chagyrskaya and Okladnikov dataset were generated from a study currently in press \cite{laurits_skov_genetic_nodate}, and the data will be uploaded to European Nucleotide Archive upon publication.
